---
title: "Olympic Games"
format: 
  dashboard:
    theme: 
      light: [sketchy, style/olympicdash.scss]
      dark: [darkly, style/olympicdash.scss]
    orientation: columns
    nav-buttons: 
      - icon: github
        href: https://github.com/posit-conf-2024/olympicdash
        aria-label: GitHub
logo: images/olympics-logo.svg
---

# Notes 🗒️

## Notes

### Row - Personal notes

**Notes**

[Speaker: Mine Cetinkays-Rundel](https://mine-cr.com/)

[GitHub - Posit Conf workshops](https://github.com/posit-conf-2024)

[GitHub - Quarto dashboards](https://github.com/posit-conf-2024/olympicdash)

[Slide](https://posit-conf-2024.github.io/quarto-dashboards/)

All levels of info interact in the `quarto dashboard` universe:

  1. yaml header;
  2. Commands such as: `# {.sidebar}`, `{.tabset}`, `{.scroll}`
  3. Levels of heading in the document:   
    i. `#` gives you a page and header button  
    ii. `##` won't give you the button, but the same level of information:
      a. Rows first: `##` = rows, `###` = columns
      b. Vice-versa
    
  4. Special characters in code chunks (e.g., `#|` + `label`, `output`, `layout-ncol`).
  
  5. Some things are set to `FALSE` automatically (e.g., `echo`, `warning`)
  
  6. Set `warning` to `TRUE` globally is a good practice to ensure that the analyses are coming out as they should. 
  
  7. We can't have `spaces` into the options, for example: `{height=10%}`, correct; vs. `{height = 10%}` wrong.  
  
*Cards* are the key structures in Quarto dashboards. First get your card for free (default behavior), but you should create/manipulated them manually for greater control. The command for manually creating the card is `:::{.card title: "text"}` and closing with `:::`. 

*Cell* options (native to Quarto) is a good way to create cards and to get automatic things done with `R` (or any language) code. For instance, using the `cat()` function to create an interactive title. 

*Value boxes* uses bootstrap icons (https://icons.getbootstrap.com), can be set to any color (there are default colors)

*Markdown text* keep it minimally, but you could add it to cell cards

*Add comments* to the script with standard `html` code `<!-- comment here -->`

*Repeat code* In the workshop, the instructor copied and pasted the content, but is there a way to avoid it? 

*Interaction* most of the time comes with a `server` element that resembles `Shiny apps`. But it is possible to include Shiny components (e.g., slide bar for filtering). Later of complexity: `Quarto dashboard > Quarto w/ Shiny components > Shiny dashboard`. When the `row/column` mentality is limited for a given purpose and interactivity is **key**, then Shiny App is a better option than Quarto dash. 

*Theming and styling* could be done with `css`, but `Quarto` can rely on [bootstrap themes](https://bootswatch.com/) that are tested and developed by a web designer. We will use the `theme` argument at the `yaml` header; you can add subthemes `dark` and `light`. You can add your own `.css` theme file. See the document `dashboard-r.qmd` for slides on fonts, background colors etc.

*Check parent <<->> child* structure as one way to automate generation of script and quarto files.

*Host shiny* there are several options for deploying dashboards with shiny/interactivy components, like [ShinyLive](https://posit-dev.github.io/r-shinylive/)


<!-- html comment: This code wont run  -->

```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(gt)
library(scales)
```

```{r}
#| label: load-data
#| message: false
olympics_full <- read_csv("data/olympics.csv")
```

```{r}
#| label: prep-data
olympics <- 
  olympics_full |>
  filter(
    #season == "Summer",
    !is.na(medal)
  ) |>
  separate_wider_delim(cols = team, names = c("team", "suffix"), delim = "-", too_many = "merge", too_few = "align_start") |>
  select(-suffix) |>
  mutate(medal = fct_relevel(medal, "Bronze", "Silver", "Gold"))
```

# Summer Olympics 🔥

## Column - Medals by sport {width=65%}

### Row - Medals by sport {height=60%}

```{r}
#| title: "Medals by sport"
#| fig-asp: 0.4
olympics |>
  filter(season == "Summer") |> 
  mutate(
    sport = fct_lump_n(sport, n = 15),
    sport = fct_infreq(sport),
    sport = fct_rev(sport),
    sport = fct_relevel(sport, "Other", after = 0)
  ) |>
  ggplot(aes(y = sport, fill = medal)) +
  geom_bar() +
  scale_fill_manual(values = c("Gold" = "#d4af37", 
                               "Silver" = "#c0c0c0",
                               "Bronze" = "#cd7f32")) +
  scale_x_continuous(breaks = seq(0, 7000, 1000),
                     labels = label_number(big.mark = ",")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Medal"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.2),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = "white", color = "gray")
  )
```

### Row - Medals by year {height=40%}

::: {.card title="Medals by year"}
Due to World War II, no Olympic games were held in 1940 and 1944.

```{r}
#| title: "Medals by year"
#| fig-width: 11
#| fig-height: 2

olympics |>
  filter(season == "Summer") |>  
  count(year, medal) |>
  ggplot(aes(x = year, y = n, color = medal)) +
  geom_point(size = 0.5) +
  geom_line() +
  scale_color_manual(values = c("Gold" = "#d4af37", 
                               "Silver" = "#c0c0c0",
                               "Bronze" = "#cd7f32")) +
  
  guides(color = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = seq(1896, 2020, 8)) +
  labs(
    x = "Year",
    y = NULL,
    color = "Medal"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.2),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = "white", color = "gray")
  )
```

:::

## Column - Medals by country {width=30%}

### Row - Value boxes {height=25%}

```{r}
#| label: country with most Gold, Silver, and Bronze medals - Summer

summer_stats <- 
  olympics |> 
  filter(season == "Summer") |> 
  group_by(medal) |> 
  # will sort the team w/ most cases on top
  count(team, sort = T) |> 
  slice_head(n = 1)
```

::: {.valuebox icon="award-fill" color="#d4af37"}
Most golds: 

`{r} summer_stats[3, 3]`

`{r} summer_stats[3, 2]`
:::

::: {.valuebox icon="award-fill" color="#c0c0c0"}
Most silvers: 

`{r} summer_stats[2, 3]`

`{r} summer_stats[2, 2]`
:::

::: {.valuebox icon="award-fill" color="#cd7f32"}
Most bronzes: 

`{r} summer_stats[1, 3]`

`{r} summer_stats[1, 2]`
:::

### Row - tabset {height=70% .tabset}

::: {.card title="Top 30 total medals"} 
Teams sorted in descending order of total medals.

```{r}
#| title: "Medals by country"
olympics |>
  filter(season == "Summer") |>  
  count(team, medal) |>
  pivot_wider(names_from = medal, values_from = n, values_fill = 0) |>
  mutate(total = Bronze + Gold + Silver) |>
  arrange(desc(total), team) |>
  slice_head(n = 30) |>
  select(Team = team, Gold, Silver, Bronze) |>
  gt() |>
  cols_align(align = "left", columns = Team) |>
  data_color(
    method = "numeric",
    palette = "nord::aurora"
  )
```
:::

::: {.card title="Bottom 30 total medals"} 
Teams sorted in descending order of total medals.

```{r}
#| title: "Medals by country"
olympics |>
  filter(season == "Summer") |>  
  count(team, medal) |>
  pivot_wider(names_from = medal, values_from = n, values_fill = 0) |>
  mutate(total = Bronze + Gold + Silver) |>
  arrange(desc(total), team) |>
  slice_tail(n = 30) |>
  select(Team = team, Gold, Silver, Bronze) |>
  gt() |>
  cols_align(align = "left", columns = Team) |>
  data_color(
    method = "numeric",
    palette = "nord::aurora"
  )
```
:::


# Winter Olympics ❄️

## Column - Medals by sport {width=65%}

### Row - Medals by sport {height=60%}

```{r}
#| title: "Medals by sport"
olympics |>
  filter(season == "Winter") |> 
  mutate(
    sport = fct_lump_n(sport, n = 15),
    sport = fct_infreq(sport),
    sport = fct_rev(sport),
    sport = fct_relevel(sport, "Other", after = 0)
  ) |>
  ggplot(aes(y = sport, fill = medal)) +
  geom_bar() +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Medal"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.2),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = "white", color = "gray")
  )
```

### Row - Medals by year {height=40%}

::: {.card title="Medals by year"}
Due to World War II, no Olympic games were held in 1940 and 1944.

```{r}
#| title: "Medals by year"
olympics |>
  filter(season == "Winter") |>  
  count(year, medal) |>
  ggplot(aes(x = year, y = n, color = medal)) +
  geom_point(size = 0.5) +
  geom_line() +
  guides(color = guide_legend(reverse = TRUE)) +
  scale_x_continuous(breaks = seq(1896, 2020, 8)) +
  labs(
    x = "Year",
    y = NULL,
    color = "Medal"
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.9, 0.2),
    legend.direction = "horizontal",
    legend.background = element_rect(fill = "white", color = "gray")
  )
```

:::

## Column - Medals by country {width=30%}

### Row - Value boxes {height=25%}

```{r}
#| label: Country with most Gold, Silver, and Bronze medals - Winter

winter_stats <- 
  olympics |> 
  filter(season == "Winter") |> 
  group_by(medal) |> 
  # will sort the team w/ most cases on top
  count(team, sort = T) |> 
  slice_head(n = 1)
```

::: {.valuebox icon="award-fill" color="#d4af37"}
Most golds: 

`{r} winter_stats[3, 3]`

`{r} winter_stats[3, 2]`
:::

::: {.valuebox icon="award-fill" color="#c0c0c0"}
Most silvers: 

`{r} winter_stats[2, 3]`

`{r} winter_stats[2, 2]`
:::

::: {.valuebox icon="award-fill" color="#cd7f32"}
Most bronzes: 

`{r} winter_stats[1, 3]`

`{r} winter_stats[1, 2]`
:::

### Row - tabset {height=70% .tabset}

::: {.card title="Top 30 total medals"} 
Teams sorted in descending order of total medals.

```{r}
#| title: "Medals by country"
olympics |>
  filter(season == "Winter") |>  
  count(team, medal) |>
  pivot_wider(names_from = medal, values_from = n, values_fill = 0) |>
  mutate(total = Bronze + Gold + Silver) |>
  arrange(desc(total), team) |>
  slice_head(n = 30) |>
  select(Team = team, Gold, Silver, Bronze) |>
  gt() |>
  cols_align(align = "left", columns = Team) |>
  data_color(
    method = "numeric",
    palette = "nord::aurora"
  )
```
:::

::: {.card title="Bottom 30 total medals"} 
Teams sorted in descending order of total medals.

```{r}
#| title: "Medals by country"
olympics |>
  filter(season == "Winter") |>  
  count(team, medal) |>
  pivot_wider(names_from = medal, values_from = n, values_fill = 0) |>
  mutate(total = Bronze + Gold + Silver) |>
  arrange(desc(total), team) |>
  slice_tail(n = 30) |>
  select(Team = team, Gold, Silver, Bronze) |>
  gt() |>
  cols_align(align = "left", columns = Team) |>
  data_color(
    method = "numeric",
    palette = "nord::aurora"
  )
```
:::

